---
title: "Analysis of lentiviral barcoding data - cDC1 versus cDC2 dendritic cells subtypes"
author: "Louisa Hadj Abed"
date: "4/06/202"
output: html_document
---

<br/>


<br/>

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(devtools)
#load packages
#devtools::install_github("TeamPerie/CellDestiny", quiet = TRUE)
library(CellDestiny)

cols = c("dark blue","#56B4E9","#0072B2",'#F15854',"#2d543d",
         "#B276B2","#CB2027",
         "#617a89","#0b53c1","#60BD68","grey90","#117733")

```


```{r, echo=TRUE, warning=FALSE, message=FALSE}
# set working directory
setwd('/Users/jasoncosgrove/Dropbox (Team_Perie)/Hadj_Abed_2022/1st_review/data_and_code/scRNA')
# import files
count_matrix <- read.csv("merged_matrices.csv")
metadata <- read.csv("merged_matrices_metadata.csv")

# Common parameters
indiv_var="mouse"
indiv_val=c("lk", "lsk")
list_var = c("type")
list_val = metadata$type
```

### Heatmap

##### Heatmap is used to visualize barcode compisition similarities between the cell types

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## Plot heatmap
distance<-"euclidean"
clusteringMeth<-"complete"
dendro<-"yes"
barcodes<-"no"
nclust<-0
pool=FALSE
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = pool)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, nClusters = nclust, columnTextSize = 1)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}


heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = TRUE)


tiff("/users/jasoncosgrove/Desktop/heatmap.tiff",units="in", width=7, height=4, res=300)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, nClusters = 6, columnTextSize = 1)
dev.off()


```


### Clone sizes

##### Plot clone-size distributions using a histogram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
indiv_val=c("lk", "lsk")
colorFreq="mouse"

freq_mat<-MakeBarcodeFrequenciesMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val)
# To have cDC2 in green :
PlotBarcodeFrequencies(freq_mat, colorFreq, y = "histogram", nbins = 10)


```

##### Plot clone-size distributions using a cumulative frequency diagram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
colorCum="type"



cum_mat<-MakeCumulativeDiagramMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val, colorCum, xProp = "yes")
p <- PlotCumulativeDiagram(cum_mat,indiv_var, colorVar = colorCum, xProp = "yes")

tiff("/users/jasoncosgrove/Desktop/cum_dist.tiff",units="in", width=8, height=4, res=300)
p + scale_color_manual(values=cols)
dev.off()
```


### Diversity 

##### Quantify clonal diversity within each cell types using the Shannon Index


```{r, echo=TRUE, warning=FALSE, message=FALSE}
boxplotColor_var=""
diversityVar<-"Shannon index"

indiv_var = "mouse"
indiv_val=c("lk" ,"lsk")

list_val<- metadata$type[1:6]

div_mat<-CalculDiversity(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val ,colorVar=boxplotColor_var, diversity=diversityVar)

library(RColorBrewer)
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(15)






tiff("/users/jasoncosgrove/Desktop/diversity.tiff",units="in", width=5, height=4, res=300)
p <-PlotDiversity(div_mat,diversityVar, list_var, indiv_var, colorVar=boxplotColor_var,textSize = 15)
p + scale_fill_manual(values=cols) + geom_point(color="black", size=2, alpha=0.9) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 


dev.off()

```

##### Now lets see if barcode abundances are similar between for example Neutrophil and Baso samples using scatter plot and pie chart visualisations

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
x_var<-c("type")
x_val<-c("Neutrophil")
y_var<-c("type")
y_val<-c("Monocyte")
color=""

indiv_val <- "lsk"
dot_mat<-MakeDotPlotMatrix(count_matrix, metadata, indiv_var, indiv_val,x_var, x_val,y_var, y_val, color)



tiff("/users/jasoncosgrove/Desktop/avsbLSK.tiff",units="in", width=5, height=5, res=300)
PlotDotplot(dot_mat, indiv_var, color, textSize = 15) 
dev.off()

pie_mat<-MakePieChartMatrix(dot_mat, indiv_var, color)
pie<-PlotPieChart(pie_mat, textSize = 5) 

# to have legend on piechart (not bellow like done by PlotPieChart function)

tiff("/users/jasoncosgrove/Desktop/pie.tiff",units="in", width=5, height=5, res=300)
pie + guides(fill = guide_legend(title = "Group")) +
    scale_y_continuous(breaks = pie_mat$ypos , labels = pie_mat$Var1)+
    theme(axis.ticks = element_blank(),
          axis.title = element_blank(),
          axis.text = element_text(size = 1), 
          legend.position = "none") + 
            scale_fill_manual(values=c(cols[8:9],"grey"))

dev.off()
```     



### Categorisation

In this section we use a threshold-based classifer approach to assign a lineage bias to each barcode. Specifically, we set thresholds of 0,10 and 20% and quantify the number of barcodes that are classified as biased.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## common parameters
# Get category variable and values
catVar<-"type"
catVal<-metadata$type[1:10]
```

##### 0% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters


catVal <- catVal[5:6]

slider<-0

mycolors <- c(cols[8:9],"grey")

count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]


tiff("/users/jasoncosgrove/Desktop/bias_0.tiff",units="in", width=5, height=5, res=300)
PlotCategories(count_per_type, slider, legendPos = "right")+ 
            scale_fill_manual(values=mycolors)+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.text=element_text(size=7), legend.title=element_blank())
dev.off()

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]

PlotCategoryCounts(count_per_cat, slider)+ 
            scale_fill_manual(values=mycolors)
dev.off()
```


##### 10% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-10


count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]

tiff("/users/jasoncosgrove/Desktop/bias_10.tiff",units="in", width=5, height=5, res=300)
PlotCategories(count_per_type, slider, legendPos = "right")+ 
            scale_fill_manual(values=mycolors)+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.text=element_text(size=7), legend.title=element_blank())
dev.off()

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]

PlotCategoryCounts(count_per_cat, slider)+ 
            scale_fill_manual(values=mycolors)

```

##### 20% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-20

count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]

tiff("/users/jasoncosgrove/Desktop/bias_20.tiff",units="in", width=5, height=5, res=300)
PlotCategories(count_per_type, slider, legendPos = "right")+ 
            scale_fill_manual(values=mycolors)+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), legend.text=element_text(size=7), legend.title=element_blank())
dev.off()

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]


PlotCategoryCounts(count_per_cat, slider)+ 
            scale_fill_manual(values=mycolors)
dev.off()
```










