---
title: "Analysis of Weinreb et al (2020)"
author: "Jason Cosgrove and Louisa Hadj Abed"
date: "4/06/202"
output: html_document
---

<br/>


<br/>

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(devtools)
library(RColorBrewer)
library(Seurat)
library(Matrix)
library(stringi)
library(stringr)
#load packages
#devtools::install_github("TeamPerie/CellDestiny", quiet = TRUE)
library(CellDestiny)

# set working directory
setwd(getwd())

cols = c("dark blue","#56B4E9","#0072B2",'#F15854',"#2d543d",
         "#B276B2","#CB2027",
         "#617a89","#0b53c1","#60BD68","grey90","#117733")

```



### Generate the count matrices and metadata

##### convert dataset into a seurat object
```{make_seurat_obj, echo=TRUE, warning=FALSE, message=FALSE}


#load in the count matrix and metadata for the weinreb et al dataset.
#these files were downloaded from https://github.com/AllonKleinLab/paper-data/tree/master/Lineage_tracing_on_transcriptional_landscapes_links_state_to_fate_during_differentiation

matrix <- readMM("stateFate_inVitro_normed_counts.mtx")
gene.names <- read.table("stateFate_inVitro_gene_names.txt",header = F)
meta.data <- read.table("stateFate_inVitro_metadata.txt",header = T)
clones <- readMM("stateFate_inVitro_clone_matrix.mtx")

#create a unique id for each celltype and update the dimnames of the matrix
meta.data$cell_id <- paste(meta.data$Cell_barcode,meta.data$Library,sep = "_")
dimnames(matrix) <- list(meta.data$cell_id,gene.names$V1)
rownames(meta.data) <- meta.data$cell_id

#update the barcode information
meta.data$clone_id <- NA

for(i in 1:ncol(clones)){
  
  #subset the metadata to get only the cells that belong to this clone
  meta.data.subset <- meta.data[meta.data$cell_id %in% names(clones[,i][clones[,i] == 1]),]
  #for each cell update the clonal information
  meta.data[meta.data$cell_id %in% names(clones[,i][clones[,i] == 1]),"clone_id"] <- i

}


#now create a seurat object
sobj <- CreateSeuratObject(counts = t(matrix), project = "weinreb", min.cells = 100, min.features = 100,meta.data = meta.data)
```



##### make the count matrix
```{make_count_matrix, echo=TRUE, warning=FALSE, message=FALSE}
lsk <- subset(sobj,Starting_population == "Lin-Kit+Sca1+")
lk <- subset(sobj,Starting_population == "Lin-Kit+Sca1-")

df.lsk <- table(lsk@meta.data$clone_id,
lsk@meta.data$Cell_type_annotation)

df.lk <- table(lk@meta.data$clone_id,
                lk@meta.data$Cell_type_annotation)


write.csv(df.lsk,"weinreb_barcodes_lsk.csv")
write.csv(df.lk,"weinreb_barcodes_lk.csv")


```

##### merge celltypes
```{r ,echo=TRUE, warning=FALSE}
lk<-read.csv("weinreb_barcodes_lk.csv")

colnames(lk)<-str_replace(colnames(lk), "_", "")
lk[,1]<-paste(lk[,1], "_lk")
rownames(lk)<-lk[,1]
lk<-lk[,-1]
colnames(lk)<-paste0(colnames(lk), "_lk")

lsk<-read.csv("weinreb_barcodes_lsk.csv")
colnames(lsk)<-str_replace(colnames(lsk), "_", "")
lsk[,1]<-paste0(lsk[,1], "_lsk")
rownames(lsk)<-lsk[,1]
lsk<-lsk[,-1]
colnames(lsk)<-paste0(colnames(lsk), "_lsk")

lk_lsk<-merge(lk, lsk, by="row.names", all=T)
lk_lsk[is.na(lk_lsk)]<-0
rownames(lk_lsk)<-lk_lsk[,1]
lk_lsk<-lk_lsk[,-1]
write.csv(file = "./merged_matrices.csv", x = lk_lsk, sep = ",")
```

```{r ,echo=TRUE, warning=FALSE}
# parameters
file_path<-"merged_matrices.csv"
field_separator=","   # if csv==,   if csv2==;   if tsv==\t
file_extention<-".csv"
sampleName_separators="_"
```

# create metadata of the given count matrix WITHOUT column names 
```{r ,echo=TRUE, warning=FALSE}
matrix<-read.csv(file_path, sep = field_separator)
# demander sample name separators
met<-MakeMetadata(matrix, sampleName_separators)
met
```

# name columns after seeing the table

```{r ,echo=TRUE, warning=FALSE}
# if not duplicats == analysis in the app
colnames(met)<-c("type", "mouse")
met
```

# save the metadata
```{r ,echo=TRUE, warning=FALSE}
file_noExtention<-gsub(pattern = file_extention, replacement = "", file_path) # change ".csv" if not .csv file
write.csv(met, paste0(file_noExtention, "_metadata.csv"), row.names = FALSE, na = "", quote = FALSE)
```

### Barcode Analysis

```{r, echo=TRUE, warning=FALSE, message=FALSE}

# import files
count_matrix <- read.csv("merged_matrices.csv")
metadata <- read.csv("merged_matrices_metadata.csv")

# Common parameters
indiv_var="mouse"
indiv_val=c("lk", "lsk")
list_var = c("type")
list_val = metadata$type
```

### Heatmap

##### Heatmap is used to visualize barcode compisition similarities between the cell types

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## Plot heatmap
distance<-"euclidean"
clusteringMeth<-"complete"
dendro<-"yes"
barcodes<-"no"
nclust<-0
pool=FALSE
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = pool)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, nClusters = nclust, columnTextSize = 0.8)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_val<-c("Neutrophil", "Monocyte")
heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = TRUE)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, columnTextSize = 1)


```


### Clone sizes

##### Plot clone-size distributions using a histogram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
indiv_val=c("lk", "lsk")
colorFreq="mouse"

freq_mat<-MakeBarcodeFrequenciesMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val)
p <- PlotBarcodeFrequencies(freq_mat, colorFreq, y = "histogram", nbins = 10)

p + scale_fill_manual(values=cols)

```

##### Plot clone-size distributions using a cumulative frequency diagram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
colorCum="type"

cum_mat<-MakeCumulativeDiagramMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val, colorCum, xProp = "yes")

p <- PlotCumulativeDiagram(cum_mat,indiv_var, colorVar = colorCum, xProp = "yes")

p + scale_fill_manual(values=cols)

```

### Diversity 

##### Quantify clonal diversity within each cell types using the Shannon Index

```{r, echo=TRUE, warning=FALSE, message=FALSE}
boxplotColor_var=""
diversityVar<-"Shannon index"

div_mat<-CalculDiversity(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val ,colorVar=boxplotColor_var, diversity=diversityVar)
p <- PlotDiversity(div_mat,diversityVar, list_var, indiv_var, colorVar=boxplotColor_var) 


p + scale_color_manual(values=cols)
```


##### Now lets see if barcode abundances are similar between for example Neutrophil and Baso samples using scatter plot and pie chart visualisations

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
x_var<-c("type")
x_val<-c("Neutrophil")
y_var<-c("type")
y_val<-c("Monocyte")
color=""

indiv_val <- "lsk"

# should not be 
dot_mat<-MakeDotPlotMatrix(count_matrix, metadata, indiv_var, indiv_val,x_var, x_val,y_var, y_val, color)
PlotDotplot(dot_mat, indiv_var, color, textSize = 15) 

pie_mat<-MakePieChartMatrix(dot_mat, indiv_var, color)
pie <- PlotPieChart(pie_mat, textSize = 5) 

pie + guides(fill = guide_legend(title = "Group")) +
    scale_y_continuous(breaks = pie_mat$ypos , labels = pie_mat$Var1)+
    theme(axis.ticks = element_blank(),
          axis.title = element_blank(),
          axis.text = element_text(size = 1), 
          legend.position = "none") + 
            scale_fill_manual(values=c(cols[8:9],"grey"))
```     

### Categorisation

In this section we use a threshold-based classifer approach to assign a lineage bias to each barcode. Specifically, we set thresholds of 0,10 and 20% and quantify the number of barcodes that are classified as biased.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## common parameters
# Get category variable and values
catVar<-"type"
catVal<- c("Monocyte","Neutrophil")

mycolors <- c(cols[8:9],"grey")

```

##### 0% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-0




count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
PlotCategories(count_per_type, slider, legendPos = "right") + scale_fill_manual(values=mycolors)


count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
PlotCategoryCounts(count_per_cat, slider) + scale_fill_manual(values=mycolors)
```


##### 10% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-10


count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
PlotCategories(count_per_type, slider, legendPos = "bottom")  + scale_fill_manual(values=mycolors)

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
PlotCategoryCounts(count_per_cat, slider) + scale_fill_manual(values=mycolors)
```

##### 20% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-20

count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
p <- PlotCategories(count_per_type, slider, legendPos = "bottom")


p + scale_fill_manual(values=mycolors) 

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
p <- PlotCategoryCounts(count_per_cat, slider)

p + scale_fill_manual(values=mycolors) 

```












