---
title: "Analysis of lentiviral barcoding data - cDC1 versus cDC2 dendritic cells subtypes"
author: "Louisa Hadj Abed"
date: "4/06/202"
output: html_document
---

<br/>


<br/>

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(devtools)
library(RColorBrewer)

#load packages
devtools::install_github("TeamPerie/CellDestiny", quiet = TRUE)
library(CellDestiny)
```


```{r, echo=TRUE, warning=FALSE, message=FALSE}
# set working directory
setwd(getwd())
# import files
count_matrix <- read.csv("merged_matrices.csv")
metadata <- read.csv("merged_matrices_metadata.csv")

# Common parameters
indiv_var="mouse"
indiv_val=c("lk", "lsk")
list_var = c("type")
list_val = metadata$type
```

### Heatmap

##### Heatmap is used to visualize barcode compisition similarities between the cell types

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## Plot heatmap
distance<-"euclidean"
clusteringMeth<-"complete"
dendro<-"yes"
barcodes<-"no"
nclust<-0
pool=FALSE
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = pool)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, nClusters = nclust, columnTextSize = 0.8)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_val<-c("Baso", "Eos")
heat_mat<-MakeHeatmapMatrix(matrix = count_matrix, metadata = metadata, indivVar = indiv_var, indivVal = indiv_val, listVar = list_var,  listVal = list_val, poolIndiv = TRUE)
p<-PlotHeatmap(heat_mat, distance = distance, clustering = clusteringMeth, showDendro = dendro, showBarcodes = barcodes, columnTextSize = 1)
```


### Clone sizes

##### Plot clone-size distributions using a histogram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
indiv_val=c("lk", "lsk")
colorFreq="mouse"

freq_mat<-MakeBarcodeFrequenciesMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val)
PlotBarcodeFrequencies(freq_mat, colorFreq, y = "histogram", nbins = 10)
```

##### Plot clone-size distributions using a cumulative frequency diagram

```{r, echo=TRUE, warning=FALSE, message=FALSE}
list_var = c("type")
list_val = metadata$type
colorCum="type"

cum_mat<-MakeCumulativeDiagramMatrix(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val, colorCum, xProp = "yes")
PlotCumulativeDiagram(cum_mat,indiv_var, colorVar = colorCum, xProp = "yes")
```

### Diversity 

##### Quantify clonal diversity within each cell types using the Shannon Index

```{r, echo=TRUE, warning=FALSE, message=FALSE}
boxplotColor_var=""
diversityVar<-"Shannon index"

div_mat<-CalculDiversity(count_matrix, metadata, indiv_var, indiv_val, list_var, list_val ,colorVar=boxplotColor_var, diversity=diversityVar)
PlotDiversity(div_mat,diversityVar, list_var, indiv_var, colorVar=boxplotColor_var)
```


##### Now lets see if barcode abundances are similar between for example Neutrophil and Baso samples using scatter plot and pie chart visualisations

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
x_var<-c("type")
x_val<-c("Neutrophil")
y_var<-c("type")
y_val<-c("Monocyte")
color=""

indiv_val <- "lsk"

# should not be 
dot_mat<-MakeDotPlotMatrix(count_matrix, metadata, indiv_var, indiv_val,x_var, x_val,y_var, y_val, color)
PlotDotplot(dot_mat, indiv_var, color, textSize = 15) 

pie_mat<-MakePieChartMatrix(dot_mat, indiv_var, color)
pie<-PlotPieChart(pie_mat, textSize = 5) 
```     

### Categorisation

In this section we use a threshold-based classifer approach to assign a lineage bias to each barcode. Specifically, we set thresholds of 0,10 and 20% and quantify the number of barcodes that are classified as biased.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## common parameters
# Get category variable and values
catVar<-"type"
catVal<-metadata$type[1:10]
```

##### 0% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-0

catVal <- catVal[5:6]

count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
PlotCategories(count_per_type, slider, legendPos = "right")


count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
PlotCategoryCounts(count_per_cat, slider)
```


##### 10% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-10


count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
PlotCategories(count_per_type, slider, legendPos = "bottom") 

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
PlotCategoryCounts(count_per_cat, slider)
```

##### 20% bias  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
## changing parameters
slider<-20

count_per_type<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[1]]
PlotCategories(count_per_type, slider, legendPos = "bottom")

count_per_cat<-MakeCategoryMatrices(count_matrix, metadata, indiv_var, indiv_val, catVar, catVal, slider)[[2]]
PlotCategoryCounts(count_per_cat, slider)
```












